\huge \textbf{Parameter Forecasts From CMB Lensing and Galaxy Lensing Power- and Bispectra}

### 2025-04-04 paper/paper.tex
fisher matrices updated
paper changes

% !TEX root = /Users/jonasfrugte/Desktop/Research_Project/paper/paper.tex
\documentclass[11pt]{article} % Choose a document class, like 'article', 'report', or 'revtex4-2' for scientific papers

@@

\affil[1]{Independent Researcher $\infty$, The Netherlands}
\affil[2]{Van Swinderen Institute for Particle Physics and Gravity, University of Groningen, Nijenborgh 4, 9747 AG Groningen, The Netherlands}
\affil[ ]{}
\affil[ ]{\textit{E-mail:} \texttt{jonasfrugte@gmail.com}}

@@

\begin{document}
\maketitle

@@

\begin{abstract}
    (Placeholder text written with chatgpt) We present a Fisher forecast analysis of the constraints on cosmological parameters from measurements of CMB lensing and galaxy weak lensing power- and bispectra. Using simulated data consistent with upcoming surveys, we explore the sensitivity of lensing observables to parameters including the Hubble constant \(H_0\), baryon density \(\Omega_b h^2\), dark matter density \(\Omega_c h^2\), scalar spectral index \(n_s\), amplitude of scalar fluctuations \(A_s\), and the sum of neutrino masses \(\sum m_\nu\).

    Our results indicate that combining power- and bispectra improves parameter constraints compared to power spectra alone, reducing uncertainties by a factor of \(\sim X\%\). In particular, we forecast constraints of \(\sigma(H_0) = X.X\) km/s/Mpc, \(\sigma(\Omega_b h^2) = X.XXX\), \(\sigma(\Omega_c h^2) = X.XXX\), \(\sigma(n_s) = X.XXXX\), \(\sigma(A_s) = X.XX \times 10^{-9}\), and \(\sigma(\sum m_\nu) = X.XX\) eV. These results demonstrate the potential of bispectrum measurements in future surveys such as the Simons Observatory, CMB-S4, LSST, and Euclid.

    Additionally, we investigate the impact of different noise models and survey redshift distributions, finding that improvements in galaxy redshift accuracy significantly enhance constraints on \(\sum m_\nu\). We also discuss the degeneracies between parameters and assess their impact using eigenvalue analysis of the Fisher matrix.

    This study highlights the advantages of combining lensing power- and bispectra for cosmological parameter estimation and provides a roadmap for optimizing future surveys to maximize their scientific return.
\end{abstract}
\pagebreak
%\tableofcontents

@@

All code written for this project is uploaded on github at LINK .

@@

Taking the harmonic decomposition of the convergence,
\begin{equation}
    \kappa(\hat{\mathbf n}) = \kappa_{lm}Y_l^m(\hat{\mathbf n}),
\end{equation}
you can relate it to the multipole moments of the lensing potential through CHECK FACTOR OF 2
\begin{equation}
    \kappa_{lm} = \frac{1}{2}(\nabla^2\psi)_{lm} = \frac{l(l+1)}{2}\psi_{lm}
\end{equation}
because the spherical harmonics are eigenfunctions of the Laplacian operator.

@@

The time complexity of the Fisher matrix scales with the maximum multipole considered, $l_{max}$, as $\mathcal O (l_{max}^3)$, leading to time issues for realistic values of $l_{max}$. We thus calculated the $l_1 < l_2 < l_3$ part of each Fisher matrix element (see appendix (REF)) by sampling every step of 10 (CHECK IF THIS IS STILL CORRECT BEFORE PUBLISHING) in $l_1$ and $l_2$. Due to the discontinuos nature of the bispectrum (e.g. it vanishes immediately if $l_1 + l_2 + l_3$ is and odd number) we didn't sample $l_3$ values. This was done similarly in (REF).

@@

For the parameter constraints we took the multipole range to be from 50 ($l_{min}$) to 1000 ($l_{max}$) (INCREASE TO 2000?) (CHECK BEFORE PUBLISHING) in line with measurement ranges from current gen surveys (PROOF BY LOOKING AT ACTUALL EXPERIMENTS).

We will consider multiple possibilities for the other calculation parameters (noise levels and redshift distribution of observed galaxies). This is done to allow us to more accurately model constraints from measurements by the Simons Observatory (SO), CMB-S4, Euclid, and Large Synoptic Survey Telescope (LSST) surveys. The first two measure CMB lensing while the latter two measure lensing shear of galaxies.

The noise levels in the CMB lensing are caused by a variety of factors, see e.g. \cite{cmblensingestimator}. The noise curves for the SO are shown alongside a simulated pure convergence power spectrum curve in figure \ref{fig:lpsplusnoise}. This is the noise source we use for CMB lensing.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/spectraplusnoise.pdf}
    \caption{CMB and galaxy lensing potential power spectra with noise. The left panel shows the galaxy lensing potential power spectrum with LSST and Euclid noise, while the right panel shows the CMB lensing potential power spectrum with OS noise.}
    \label{fig:lpsplusnoise}
\end{figure}

For shear measurements, the noise is simpler to model as the primary source is shot noise due to the discrete number of galaxies in each patch of the survey. The mean ellipiticity of the galaxies in a patch is calculated and a deviation from the mean is interpreted as lensing shear. The greater the number of galaxies the more accurately can the mean ellipticity be determined. Choosing the size of the patch is a balancing between signal to noise and resolution. Smaller patches give more uncertain shear measurements while providing better overal resolution for the shear map and thus allowing measurements of shear bi- and powerspectra to higher multipole numbers. The opposite is true for larger patch sizes. The power spectrum of the shear noise is given by
\begin{equation*}
    N^{\gamma\gamma}_{l} = \frac{\sigma^2_{rms}}{\bar n_g}
\end{equation*}
where $\bar n_g$ is the mean number of galaxies observed per sterradian accross the whole survey \cite{weaklensingnotes2020}.
and $\sigma^2_\gamma$ is the intrinsic ellipticity dispersion. 
Euclid and the LSST both have $\sigma_{rms} \approx 0.26$ \cite{euclid1overview} \cite{lsstsciencebook}.
Euclid is expected to measure a galaxy density of approximately 30 galaxies per square arcminute \cite{euclidprep1}
while LSST aims to measure 26 \cite{galaxydenslsst}
The noise levels are graphed alongside a simulated shear power spectrum in figure \ref{fig:lpsplusnoise}. We perform our calculations with Euclid noise levels throughout this paper.
The redshift distribution of the observed galaxies is commonly parameterized as \cite{bartelmann2001weak}
\begin{equation*}
    n(z) \propto z^a \exp\sbr{-\br{\frac{z}{z_0}}^b}.
\end{equation*}
We choose the parameter combination $a = 2$, $b = 3/2$ and $z_0 = 0.64$ which is similar to the expected distributions of Euclid, (which will probe primarily in the 0.2 - 2.6 redshift range \cite{euclidprep10})
and the LSST mission (which has $a \approx 2$, $b \approx 1$, and $z_0 \approx 0.3$ from predictions for the obtained data \cite{lsstsciencebookchapter3}).

\section{Results}\label{sec:results}
\subsection{SNR} 
The signal to noise ratios of the CMB and galaxy lensing power- and bispectra v.s. the maximum multipole moment measured can be found in figure \ref{fig:snrplots}. It can be seen that the CMB bispectrum signal to noise never exceeds 2 even for $l_{\max} = 2000$, meaning that this signal is not likely to be able to contribute to parameter constraints in future surveys. All other quantities should be well detectable for $l_{\max}\geq 500$.
\begin{figure}
    \includegraphics[width=\textwidth]{figures/snrplots.pdf}
    \caption{Signal to noise ratios for galaxy lensing (dashed) and CMB lensing (non-dashes) power- (in red) and bispectra (in blue). The $x$-axis labels the maximum multipole moment measured up to, starting from $l_{\min}=50$.}
    \label{fig:snrplots}
\end{figure}

>>>>

% !TEX root = /Users/jonasfrugte/Desktop/Research_Project/fisher_calc_weak_lensing/paper/paper.tex
\documentclass[11pt]{article} % Choose a document class, like 'article', 'report', or 'revtex4-2' for scientific papers

@@

\affil[1]{Independent Researcher, The Netherlands}
\affil[2]{Van Swinderen Institute for Particle Physics and Gravity, University of Groningen, Nijenborgh 4, 9747 AG Groningen, The Netherlands}
\affil[ ]{}
\affil[ ]{\textit{E-mail:} \texttt{jonasfrugte@gmail.com}}

@@

\begin{document}
\begin{center}
    \textit{Please note:}\\ This is still a draft version of the paper I am planning to finish April 2025. It thus may still contain mistakes, missing citations, typo's, etc. that will not be present in the final version. The code used to write this paper can be found on my GitHub: \url{https://github.com/Jonas-Frugte/fisher_calc_weak_lensing}.
\end{center}
\pagebreak
\maketitle

@@

\begin{abstract}
    Weak gravitational lensing of the Cosmic Microwave Background (CMB) and distant galaxies provides a powerful probe of the large-scale structure and cosmic expansion history. Upcoming Stage-IV surveys like the Simons Observatory, CMB-S4, LSST, and Euclid promise unprecedented precision, motivating detailed forecasts of their constraining power. While lensing power spectra are standard cosmological tools, non-Gaussianities induced by structure formation encode additional information in the bispectrum. In this work, we perform a comprehensive Fisher matrix forecast to quantify the detectability of these signals and the cosmological constraints achievable by combining power spectrum and bispectrum measurements from both CMB lensing and galaxy shear, including their cross-correlations. We model Stage-IV-like experimental specifications and forecast constraints on the standard $\Lambda$CDM parameters, the sum of neutrino masses ($\Sigma m_\nu$), and the dark energy equation of state ($w_0$). We demonstrate that incorporating bispectra significantly enhances constraints, reducing marginalized parameter uncertainties by up to $\sim \texttt{X}\%$ compared to power spectra alone. We project competitive constraints, including $\sigma(H_0) = \texttt{X.X~km/s/Mpc}$, $\sigma(\sigma_8) = \texttt{X.XXX}$, and $\sigma(\Sigma m_\nu) = \texttt{X.XX~eV}$.
    \end{abstract}
\pagebreak
%\tableofcontents

@@

All code written for this project is uploaded on github at \url{https://github.com/Jonas-Frugte/fisher_calc_weak_lensing}.

@@

Taking the harmonic decomposition of the convergence,
\begin{equation}
    \kappa(\hat{\mathbf n}) = \kappa_{lm}Y_l^m(\hat{\mathbf n}),
\end{equation}
you can relate it to the multipole moments of the lensing potential through
\begin{equation}
    \kappa_{lm} = \frac{1}{2}(\nabla^2\psi)_{lm} = \frac{l(l+1)}{2}\psi_{lm}
\end{equation}
because the spherical harmonics are eigenfunctions of the Laplacian operator.

@@

The time complexity of the Fisher matrix scales with the maximum multipole considered, $l_{max}$, as $\mathcal O (l_{max}^3)$, leading to time issues for realistic values of $l_{max}$. We thus calculated the $l_1 < l_2 < l_3$ part of each Fisher matrix element (see the appendix) by sampling every step of 10 %(CHECK IF THIS IS STILL CORRECT BEFORE PUBLISHING) 
in $l_1$ and $l_2$. Due to the discontinuos nature of the bispectrum (e.g. it vanishes immediately if $l_1 + l_2 + l_3$ is and odd number) we didn't sample $l_3$ values. This was done similarly in (REF).

@@

For the parameter constraints we took the multipole range to be from 2 ($l_{min}$) to 2000 ($l_{max}$).

% We will consider multiple possibilities for the other calculation parameters (noise levels and redshift distribution of observed galaxies). This is done to allow us to more accurately model constraints from measurements by the Simons Observatory (SO), CMB-S4, Euclid, and Large Synoptic Survey Telescope (LSST) surveys. The first two measure CMB lensing while the latter two measure lensing shear of galaxies.

For noise in CMB lensing we consider lensing estimated by an experiment measuring polarization with white noise $\Delta_P$ and beam spread $\sigma$. We then use the quadratic estimator from \cite{Hu_2002} to estimate the resulting noise in the lensing power spectrum.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{/Users/jonasfrugte/Desktop/Research_Project/fisher_calc_weak_lensing/code/plots/spectraplusnoise.pdf}
    \caption{CMB (right) and galaxy (left) lensing potential power spectra compared to associated experimental noise. Current (stage 3) noise values are displayed as well as near future (stage 4) noi se values. The CMB lensing experiment does not measure temperature anisotropies, only polarization. CMB noise values are chosen in accordance with \cite{Namikawa_2016}. Shear noise values are chosen to be similar to e.g. Euclid measurements for stage 4 and e.g. KiDS for stage 3.}
    \label{fig:lpsplusnoise}
\end{figure}

For shear measurements, the noise is simpler to model as the primary source is shot noise due to the discrete number of galaxies in each patch of the survey. The mean ellipiticity of the galaxies in a patch is calculated and a deviation from the mean is interpreted as lensing shear. The greater the number of galaxies the more accurately can the mean ellipticity be determined. Choosing the size of the patch is a balancing between signal to noise and resolution. Smaller patches give more uncertain shear measurements while providing better overal resolution for the shear map and thus allowing measurements of shear bi- and powerspectra to higher multipole numbers. The opposite is true for larger patch sizes. The power spectrum of the shear noise is given by
\begin{equation*}
    N^{\gamma\gamma}_{l} = \frac{\sigma^2_{rms}}{\bar n_g}
\end{equation*}
where $\bar n_g$ is the mean number of galaxies observed per sterradian accross the whole survey \cite{weaklensingnotes2020}.
and $\sigma^2_\gamma$ is the intrinsic ellipticity dispersion. 
Experiments such as Euclid and the LSST both have $\sigma_{rms} \approx 0.3$ \cite{euclid1overview} \cite{lsstsciencebook} and 
$ n_g\approx 30 \text{ arcmin}^{-2}$.

The noise values are calculated for stage 3 (current) surveys and stage 4 (near future) surveys. We will use the stage 4 noise values for parameter constraints. A comparison of stage 3 noise, stage 4 noise, and lensing signal can be seen in figure \ref{fig:lpsplusnoise}.

The redshift distribution of the observed galaxies is commonly parameterized as \cite{bartelmann2001weak}
\begin{equation*}
    n(z) \propto z^a \exp\sbr{-\br{\frac{z}{z_0}}^b}.
\end{equation*}
We choose the parameter combination $a = 2$, $b = 3/2$ and $z_0 = 0.64$ which is similar to the expected distributions of Euclid, (which will probe primarily in the 0.2 - 2.6 redshift range \cite{euclidprep10})
and the LSST mission (which has $a \approx 2$, $b \approx 1$, and $z_0 \approx 0.3$ from predictions for the obtained data \cite{lsstsciencebookchapter3}).

\section{Results}\label{sec:results}
\subsection{SNR}
Lensing power spectra are well detectable in nearly all cases. 
The signal to noise ratios of the CMB and galaxy lensing bispectra v.s. the maximum multipole moment measured can be found in figure \ref{fig:snrplots}. Shear bispectra can be measured well by both stage 3 and stage 4 experiments, while CMB lensing is only semi detectable by a stage 3 experiment but well detectable by stage 4 experiments.
%It can be seen that the CMB bispectrum signal to noise never exceeds 2 even for $l_{\max} = 2000$, meaning that this signal is not likely to be able to contribute to parameter constraints in future surveys. All other quantities should be well detectable for $l_{\max}\geq 500$.
\begin{figure}
    \includegraphics[width=\textwidth]{figures/snrplots.pdf}
    \caption{Signal to noise ratios for galaxy lensing (dashed) and CMB lensing (non-dashes) power- (in red) and bispectra (in blue). The $x$-axis labels the maximum multipole moment measured up to, starting from $l_{\min}=2$.}
    \label{fig:snrplots}
\end{figure}
